---
title: 'Loading lab data: first exploration'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## 1. Loading lab data: first exploration

In this section, we outline the initial steps taken to load and inspect the lab data set. The primary goal is to understand the structure of the data and to translate column names from German to English.

### 1.1 Load the data and necessary packages

```{r}
# Load Packages
library(data.table)

# Input file path for the lab data
lab_data_path <- "./RITM0154633_del_20240923/RITM0154633_lab.csv"
lab_data <- read.csv(lab_data_path)
```

### 1.2 Basic data inspection

-   Check the first rows to see the structure of the data set:

```{r}
head(lab_data)
```

*The data set contains columns such as patient and case pseudonyms, test names, abbreviations, and numerical results.*

-   Print column names:

```{r}
names(lab_data)
```

-   Since the data set contains German terms, translate columns for non-German speakers:

```{r}
colnames(lab_data) <- c("pseudo_patient_id", "pseudo_case_id", "test_name", 
                         "test_abbr", "method_num", "num_result", 
                         "text_result", "unit")
```

### 1.3 Data dimensions and overview

-   Get the dimensions, column names (after translation), structure and summary statistics:

```{r}
# Number of rows and columns
dim(lab_data)
```

The lab_data data set consists of 20,487,004 observations and 8 variables.

```{r}
# Column names
names(lab_data)
```

```{r}
# View structure of the dataset
str(lab_data)
```

```{r}
# Get summary statistics for all columns
summary(lab_data)
```

### 1.4 Additional data checks

-   Check for missing values:

```{r}
missing_values <- sapply(lab_data, function(x) sum(is.na(x)))
print(missing_values)
```

*It has **0** missing values for most columns, but the `test_abbr` column has **387,136** missing entries, and the `text_result` column has **39** missing values.*

\
- Get unique values for all columns and count them:

```{r}
unique_values_list <- lapply(lab_data, unique)
unique_counts <- sapply(unique_values_list, length)
unique_df <- data.frame(
  Unique_Count = unique_counts,
  Unique_Values = I(unique_values_list)  # Using I() to prevent auto-unlisting
)
print(unique_df)
```

-   Get unique test abbreviations, sort them, and display the result:

```{r}
unique_test_abbr <- sort(unique(lab_data$test_abbr))
unique_abbr_df <- data.frame(Unique_Test_Abbreviations = unique_test_abbr)
head(unique_abbr_df)
```

-   Get unique test names, sort them, and display the result:

```{r}
unique_test_names <- sort(unique(lab_data$test_name))

unique_names_df <- data.frame(Unique_Test_Names = unique_test_names)

head(unique_names_df)
```

-   Create a list of unique pairs of test names and their abbreviations from the `lab_data` data frame. These pairs are then sorted by the `test_name` for easier viewing and analysis. Finally, look at the first few rows to verify that it's correct:

```{r}
# Create unique test_name and test_abbr pairs
# Get unique pairs of test_name and test_abbr
unique_test_pairs <- unique(lab_data[, c("test_name", "test_abbr")])

# Sort the pairs by test_name for better readability
unique_test_pairs_sorted <- unique_test_pairs[order(unique_test_pairs$test_name), ]

# View the first few rows of the sorted unique test pairs
head(unique_test_pairs_sorted)
```

-   Check for duplicate rows in the data set:

```{r}
lab_data_dt <- as.data.table(lab_data) # Convert to data.table

# Find the duplicate rows
duplicates <- lab_data_dt[duplicated(lab_data_dt) | duplicated(lab_data_dt, fromLast = TRUE)]

# Count the number of duplicates
duplicates_count <- nrow(duplicates)
print(paste("Number of duplicate rows:", duplicates_count))

# Print the duplicate rows
print(duplicates)
```

*There are 25 duplicate rows identified.*

### 1.5 Save any initial inspection results

```{r eval=FALSE, include=TRUE}
# Save the lab data with English column names
write.csv(lab_data, "output_files/lab_data_eng_cols.csv", row.names = FALSE)

# Save column names in a CSV file
# write.csv(names(lab_data), "lab_data_column_names.csv", row.names = FALSE)

# Save unique_test_abbreviations in a CSV file
write.csv(unique_abbr_df, "output_files/unique_test_abbreviations.csv", row.names = FALSE)

# Save unique_test_names in a CSV file
write.csv(unique_names_df, "output_files/unique_test_names.csv", row.names = FALSE)

# Save unique test_name and test_abbr pairs to a CSV file
write.csv(unique_test_pairs_sorted, "output_files/unique_test_name_abbr_pairs.csv", row.names = FALSE)
```

### Data set Summary (Final Overview):

*The **lab data** dataset contains **20,487,004 observations** and **8 variables**, with a file size of **1.0 GB**. Most columns have **0 missing values**, but there are some exceptions:*

-   *The `test_abbr` column has **387,136 missing entries**.*

-   *The `text_result` column has **39 missing values**.*

*Additionally:*

-   ***25 duplicate rows** were identified within the data set.*

-   *The data set includes **182,238 unique patients** and **311,611 unique case identifiers**.*

-   *There are **3,470 unique test names** and **4,609 unique test abbreviations**.*

#### Variable Descriptions:

| **Variable Name** | **Data Type** | **Description** |
|------------------------|------------------------|------------------------|
| `pseudo_patient_id` | Integer | Unique identifier for each patient. |
| `pseudo_case_id` | Integer | Unique identifier for each case. |
| `test_name` | Character | Name of the lab test. |
| `test_abbr` | Character | Abbreviation of the lab test (if available). |
| `method_num` | Integer | Numerical identifier for the testing method. |
| `num_result` | Character | Numeric result of the lab test. |
| `text_result` | Character | Textual representation of the test result. |
| `unit` | Character | Unit of measurement for the result. |
